Nonlinear mixed effects models for time series forecasting of smart meter demand

### 2020-09-30 paper.Rmd
Fix typo
Change one-day to one day.

---
title: "Nonlinear mixed effects models for time series forecasting of smart meter demand"
author:
- familyname: Roach
  othernames: Cameron
  address: Monash University
  email: cameron.roach@monash.edu
  correspondingauthor: true
- familyname: Hyndman
  othernames: Rob
  address: Monash University
- familyname: Taieb
  othernames: Souhaib Ben
  address: University of Mons
abstract: Buildings are typically equipped with smart meters to measure electricity demand at regular intervals. Smart meter data for a single building have many uses, such as forecasting and assessing overall building performance. However, when data are available from multiple buildings, there are additional applications that are rarely explored. For instance, we can explore how different building characteristics influence energy demand. If each building is treated as a random effect and building characteristics are handled as fixed effects, a mixed effects model can be used to estimate how characteristics affect energy usage. In this paper we demonstrate that producing one-day ahead demand predictions for 123 commercial office buildings using mixed models can improve forecasting accuracy. We experiment with random intercept, random intercept and slope, and nonlinear mixed models. The predictive performance of the mixed effects models are tested against naive, linear and nonlinear benchmark models fitted to each building separately. This research justifies using mixed models to improve forecasting accuracy and to quantify changes in energy consumption under different building configuration scenarios.
keywords: "time series forecasting, mixed-effects models, smart meters, energy, electricity"
wpnumber: 1/19
jelcodes: C10,C14,C52
blind: false
cover: false
toc: false
bibliography: ["library.bib", "packages.bib"]
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

@@

This paper explores how electricity forecasting accuracy can be improved by using mixed effects models. We examine if mixed models can produce forecasts as accurately as separate models fit for each subject. We approach the problem in the context of producing one-day ahead forecasts of electricity demand for `r n_buildings` commercial office buildings in Australia. Electricity demand readings at 15-minute intervals are taken from smart meters located at each of the buildings. These readings exhibit within-subject correlation which are well suited to modelling using mixed effects models, which we explore here. When working with mixed effects models, each building is treated as a random effect and building characteristics are treated as fixed effects. We attempt to model the relationship between temperature and demand using both linear and spline based methods.

@@

Training data are comprised of business days within a window of `r fcst_train_window` days prior to the month we wish to select variables for. As our experimental setup for the validation phase involves one-day ahead forecasts, this variable selection prevents us from using any data from the future. For example, when forecasting for any date in January, only data from the months preceding January would have been used to select predictors.

@@

In order to build a better understanding of how a mixed model framework improves upon fitting individual models to each building we need to assess each model's performance. To do so, we create one-day ahead ex-post forecasts and calculate the MAE, MAPE, sMAPE and MASE for each. We focus on ex-post forecasting as we wish to examine error caused by model specification and ignore errors caused by incorrect weather forecasts, as would be the case in an ex-ante forecasting scenario.

@@

(ref:one-day-ahead-forecast-cap) One-day ahead forecasts for `r format(p_one_day_ahead_date, "%e %B, %Y")`. The Naive (yesterday's actuals) and SSC models are shown. The naive model often includes erratic spikes whereas the SSC model tends to produce a smoother profile.

@@

We fit models to `r n_buildings` buildings across Australia. Separate models for each building were fitted as a benchmark. The overall predictive power of several mixed effects models were assessed against this benchmark. One-day ahead forecasts were produced for business days over a year using all forecast methods. Based on the MAE, MAPE, sMAPE and MASE scores of each model the SSC and SSCATTR models performed best. We concluded that predicting electricity demand using nonlinear mixed effects models can improve forecast accuracy.

>>>>

---
title: "Nonlinear mixed effects models for time series forecasting of smart meter demand"
author:
- familyname: Roach
  othernames: Cameron
  address: Monash University
  email: cameron.roach@monash.edu
  correspondingauthor: true
- familyname: Hyndman
  othernames: Rob
  address: Monash University
- familyname: Taieb
  othernames: Souhaib Ben
  address: University of Mons
abstract: Buildings are typically equipped with smart meters to measure electricity demand at regular intervals. Smart meter data for a single building have many uses, such as forecasting and assessing overall building performance. However, when data are available from multiple buildings, there are additional applications that are rarely explored. For instance, we can explore how different building characteristics influence energy demand. If each building is treated as a random effect and building characteristics are handled as fixed effects, a mixed effects model can be used to estimate how characteristics affect energy usage. In this paper we demonstrate that producing one day ahead demand predictions for 123 commercial office buildings using mixed models can improve forecasting accuracy. We experiment with random intercept, random intercept and slope, and nonlinear mixed models. The predictive performance of the mixed effects models are tested against naive, linear and nonlinear benchmark models fitted to each building separately. This research justifies using mixed models to improve forecasting accuracy and to quantify changes in energy consumption under different building configuration scenarios.
keywords: "time series forecasting, mixed-effects models, smart meters, energy, electricity"
wpnumber: 1/19
jelcodes: C10,C14,C52
blind: false
cover: false
toc: false
bibliography: ["library.bib", "packages.bib"]
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

@@

This paper explores how electricity forecasting accuracy can be improved by using mixed effects models. We examine if mixed models can produce forecasts as accurately as separate models fit for each subject. We approach the problem in the context of producing one day ahead forecasts of electricity demand for `r n_buildings` commercial office buildings in Australia. Electricity demand readings at 15-minute intervals are taken from smart meters located at each of the buildings. These readings exhibit within-subject correlation which are well suited to modelling using mixed effects models, which we explore here. When working with mixed effects models, each building is treated as a random effect and building characteristics are treated as fixed effects. We attempt to model the relationship between temperature and demand using both linear and spline based methods.

@@

Training data are comprised of business days within a window of `r fcst_train_window` days prior to the month we wish to select variables for. As our experimental setup for the validation phase involves one day ahead forecasts, this variable selection prevents us from using any data from the future. For example, when forecasting for any date in January, only data from the months preceding January would have been used to select predictors.

@@

In order to build a better understanding of how a mixed model framework improves upon fitting individual models to each building we need to assess each model's performance. To do so, we create one day ahead ex-post forecasts and calculate the MAE, MAPE, sMAPE and MASE for each. We focus on ex-post forecasting as we wish to examine error caused by model specification and ignore errors caused by incorrect weather forecasts, as would be the case in an ex-ante forecasting scenario.

@@

(ref:one-day-ahead-forecast-cap) One day ahead forecasts for `r format(p_one_day_ahead_date, "%e %B, %Y")`. The Naive (yesterday's actuals) and SSC models are shown. The naive model often includes erratic spikes whereas the SSC model tends to produce a smoother profile.

@@

We fit models to `r n_buildings` buildings across Australia. Separate models for each building were fitted as a benchmark. The overall predictive power of several mixed effects models were assessed against this benchmark. One day ahead forecasts were produced for business days over a year using all forecast methods. Based on the MAE, MAPE, sMAPE and MASE scores of each model the SSC and SSCATTR models performed best. We concluded that predicting electricity demand using nonlinear mixed effects models can improve forecast accuracy.
